The \project{unRAVE} catalog

### 2016-09-04 article/unrave.tex
Updates

The target selection for \rave\ is based on the $I$-band apparent magnitude,
$9 < I < 12$, with a weak $J - K_s > 0.5$ cut near the disk and bulge \citep{Wonjo_2016}.  
The $I$ band was used for the target selection because it has a good overlap with the
wavelength range that \rave\ operates in:  8410--8795~\AA.  This spectral region 
is very well-studied.  It is one of the key setups used for the ground-based 
high-resolution \ges\ survey \citep{Gilmore_2012,Randich_2013}, and the 
\rave\ resolution and wavelength coverage is comparable to the Radial Velocity 
Spectrometer on board the \gaia\ space telescope \citep{Kordopatis_2011,Recio-Blanco_2016}.  
The region includes the \ion{Ca}{2} near infrared triplet lines --- strong 
transitions that are dominated by pressure broadening --- which are visible even 
in very low signal-to-noise (S/N) data or metal-poor stars.  Atomic transitions of 
light-, $\alpha$-, and Fe-peak elements are also present, allowing for detailed 
chemical abundance studies.

@@

The red giant branch stars in our training set have have stellar parameters 
($\teff$, $\logg$) and up to 15 individual abundances from the \aspcap\ 
\citep{aspcap_paper}.  A subset of these elements have atomic transitions in the 
\rave\ wavelength region: \ion{O}{1}, \ion{Mg}{1}, \ion{Al}{1}, \ion{Si}{1}, 
\ion{Ca}{2}, \ion{Ti}{1}, \ion{Fe}{1}, and \ion{Ni}{1}.  However, we 
excluded [Ti/H] from our abundance list because of systematics in the training
set \citep{Holtzman_2015,Hawkins_2016}.  Therefore we are left with nine labels 
in our giant star model: $\teff$, $\logg$, and seven elemental abundances.

@@

% For each spectrum we have results from three different models, described in the previous sections

% the artefacts show that the models poorly extrapolate outside the training set.
% the joint model shows that a single quadratic model is too simple to apply to the entire rave data

% we cannot just use reduced chi sq to exclude stars, nor can we just exclude things that are
% outside the range of values in the training set --> stars project back into the label space

% therefore we will use the joint model to inform us as to whether we should adopt
% results from the giant star model, the dwarf model, or a linear combination of the two.

% Show results compared to the joint model.
% Central density of points that agree well with the joint model.
% Sigmas chosen to scale both models so that the central densities are approximately isotropic.
% The artefacts in the previous figures appear here too, in strange locations.

% Equations for calculating weights

% Summing bsaed on weights

% Show HRD with mean weight fraction

\begin{figure*}[t!]
\includegraphics[width=\textwidth]{figures/test-set-density.pdf}
\caption{\label{fig:test-set-density}}
\end{figure*}

\begin{figure}[h!]
\includegraphics[width=0.5\textwidth]{figures/joint-model-differences.pdf}
\caption{\label{fig:joint-model-differences}}
\end{figure}

@@

In addition to the cross-validation tests that we have previously described, 
we conducted a number of internal and external validation experiments to 
test the validity of our results.

@@

\stub{Covariances between these parameters are accessible through the matrix of correlation coefficients in the associated results file.}

\subsection{Quality control}
\stub{What should we adopt as recommended QC? Show rogue gallery of a few "high s/n cases" where things went haywire?}

@@

Here we cross-match our results against literature samples to verify that
our labels are accurate.  Because this work is an independent analysis, our
first point of reference is against the official \rave\ data releases.  These
comparisons are shown in Figures \ref{fig:rave-dr4-comparison} and 
\ref{fig:rave-dr5-comparison}, where we show our labels with respect to the
results in the fourth and fifth \rave\ data releases, respectively.  In order
to provide a fair comparison, we only show stars that meet a number of quality
flags in both samples. 
\stub{What selections qualify as a fair comparison in both samples?}
There is very good agreement in $\teff$, with a bias
and RMS of just XX~K and XXX~K, respectively.  We find our $\logg$ values are
closest to the `calibrated' \rave\ surface gravities, which make use of 
asteroseismic information to correct (increase) the $\logg$ values along the
giant branch.  The calibrated \rave\ surface gravities also correct the 
location of the red clump (by $\sim0.5$~dex) to be in the same position in our
results.  However, while our $\logg$ values generally agree with the calibrated
\rave\ values along the giant branch, our $\logg$ values do differ slightly 
along the main-sequence.  This is principally because the main-sequence in 
this work tends to taper down towards higher $\logg$ values at cooler 
temperatures, whereas the comparison sample tend to have a slightly flatter
main-sequence.  This difference is not likely to have a very significant 
effect on the detailed abundance or spectrophotometric distance determinations
between these studies \citep{Binney_2014}.

@@

\stub{Any differences in RAVE abundances? Corrections required due to having an unbalanced training set? Any caution required?}

@@

After verifying that our stellar labels are comparable with other
high-resolution studies, here we verify that the abundance labels we find
are consistent with expectations from astrophysics.  One example is stars in
globular clusters, which show anti-correlations in light element abundances 
\citep[e.g.,][and references therein]{Da_Costa,Carretta_2009}. In the \rave\
survey, \cite{Anguiano_2015} found 70 stars with positions and radial velocities that are
consistent with being members of globular clusters.  We have used that sample
Specifically, this GC sample includes 49 stars belonging to NGC~5139 ($\omega$~Centauri), 11 members of the retrograde
globular cluster NGC~3201, and 10 members of NGC~362.  
All of these systems are well-studied and exhibit abundance anti-correlations 
in Na--O and Mg--Al \citep{people}.  Other correlations are also present 
(e.g., C--N), but those relationships are less relevant for this work because 
we do not measure C and N abundances here.  We note that \emph{none} of these 
cluster stars are present in our training set.

@@

\acknowledgements
We thank Jonathan Bird (Vanderbilt), Sven Buder (MPIA), and Sergey Koposov (Cambridge).
This research made use of: 
  Astropy, a community-developed core Python package for Astronomy \citep{astropy};
  NASA's Astrophysics Data System Bibliographic Services;
% numpy, scipy, matplotlib.
and 
  \project{TOPCAT} \citep{Taylor_2005}.
This work was partly supported by the European Union FP7 programme through ERC 
grant number 320360. Funding for K.~H.~ has been provided through the Simons 
Foundation Society of Fellows and the Marshall Scholarship.
Funding for RAVE has been provided by: the Australian Astronomical Observatory; 
the Leibniz-Institut fuer Astrophysik Potsdam (AIP); the Australian National 
University; the Australian Research Council; the French National Research Agency;
the German Research Foundation (SPP 1177 and SFB 881); the European Research 
Council (ERC-StG 240271 Galactica); the Istituto Nazionale di Astrofisica at 
Padova; The Johns Hopkins University; the National Science Foundation of the USA
(AST-0908326); the W. M. Keck foundation; the Macquarie University; the 
Netherlands Research School for Astronomy; the Natural Sciences and Engineering 
Research Council of Canada; the Slovenian Research Agency; the Swiss National 
Science Foundation; the Science \& Technology Facilities Council of the UK; 
Opticon; Strasbourg Observatory; and the Universities of Groningen, Heidelberg 
and Sydney. The RAVE web site is https://www.rave-survey.org.

@@

@@

@@

@@

@@

@@

@@

@@

@@

>>>>

The target selection for \rave\ is based on the $I$-band apparent magnitude,
$9 < I < 12$, with a weak $J - K_s > 0.5$ cut near the disk and bulge \citep{Wonjo_2016}.  
The $I$ band was used for the target selection because it has a good overlap with the
wavelength range that \rave\ operates in:  8410--8795~\AA.  This spectral region 
is very well-studied.  It is one of the key setups used for the ground-based 
high-resolution \ges\ survey \citep{Gilmore_2012,Randich_2013}, and the 
\rave\ resolution and wavelength coverage is comparable to the Radial Velocity 
Spectrometer on board the \gaia\ space telescope \citep{Kordopatis_2011,Recio-Blanco_2016}.  
The region includes the \ion{Ca}{2} near infrared triplet lines --- strong 
transitions that are dominated by pressure broadening --- which are visible even 
in metal-poor stars or spectra with very low signal-to-noise (S/N) ratios.
Atomic transitions of 
light-, $\alpha$-, and Fe-peak elements are also present, allowing for detailed 
chemical abundance studies.

@@

The red giant branch stars in our training set have have stellar parameters 
($\teff$, $\logg$) and up to 15 elemental abundances from the \aspcap\ 
\citep{Garcia_Perez_2016}.  A subset of these elements have atomic transitions in the 
\rave\ wavelength region: \ion{O}{1}, \ion{Mg}{1}, \ion{Al}{1}, \ion{Si}{1}, 
\ion{Ca}{2}, \ion{Ti}{1}, \ion{Fe}{1}, and \ion{Ni}{1}.  However, we 
excluded [Ti/H] from our abundance list because of systematics in the \aspcap\
[Ti/H] abundances \citep{Holtzman_2015,Hawkins_2016}.  Therefore we are left 
with nine labels in our giant star model: $\teff$, $\logg$, and seven elemental 
abundances.

@@

We have derived labels for all \Nspectra\ \rave\ spectra using the three models
described in previous sections.  The results from the joint model --- that includes 
the main-sequence, sub-giant and red giant branch --- shows that a single quadratic
model is too simple for the \rave\ spectral range.  The other models have problems,
too: unrealistic over-densities in label space show that the main-sequence model and
the giant model make very poor extrapolations for stars outside their respective
training sets.  For these reasons we were forced to exclude or severely penalize 
incorrect results from both models.

Before attempting to join the results from different models, we excluded results
in either model that had a reduced $\chi_{r}^2 > 3$.  We further discarded stars with
labels that are outside the extent of the training set.  Specifically for the
results from the giant model we (conservatively) excluded stars with derived 
$\logg > 3.5$, and for the results from the main-sequence model we excluded 
sub-giant stars ($\logg < 4$ and $\teff < 5000$~K) that were outside the 
two-dimensional ($\teff$, $\logg$) convex hull of the training set used for the main-sequence model.  Unfortunately these restrictions did not remove all spurious
results.  The reason for this can be explained with an example:  consider that our 
giant star model was trained with only giant stars but tested with both giant 
stars and dwarf stars.  Some classes of stars (e.g., metal-poor dwarfs) can 
project into a region of label space that would suggest it is a giant 
(e.g., a clump star).  These objects could have relatively low $\chi_{r}^2$ values
(e.g., $\chi_{r}^2 < 3$) and in this example, they would appear as bonafide red clump 
stars.  These incorrect projections are extrapolation errors in high dimensions that
project to `normal' parts of the label space in two dimensions.  For these reasons 
we also made use of the joint model in Section~\ref{sec:a-simple-model}
to inform whether we should adopt results from: the red giant branch model; the
main-sequence/sub-giant model; or a linear combination of the two.

In Figure \ref{fig:joint-model-differences} we show the differences in effective temperature
$\teff$ and surface gravity $\logg$ between: the main-sequence model and the joint
model; and the differences between the red giant branch model and the joint model. We have scaled
the differences in $\teff$ and $\logg$ to make the central peak near $(0, 0)$ to 
% ARC to HOGG: This is a total hack. Should we fit for these and \rho_{\teff,\logg} or are you content with the hack?
be approximately isotropic by setting
	$\delta_{T_{\rm eff}} = 90$~K for the main-sequence model, 
	$\delta_{T_{\rm eff}} = 50$~K for the giant model, and 
	$\delta_{\log{g}} = 0.15$~dex for both models.
The stars within the peak at $(0, 0)$ represent objects where
the joint model and the comparison model both report similar labels.  The artefacts 
seen in the Hertszprung-Russell diagrams in Figure \ref{fig:test-set-density} are
also present in Figure \ref{fig:joint-model-differences} as over-densities far away
from the central peak.  Therefore, we can 
adopt the scaled distance in labels $\teff$ and $\logg$ from the joint model,
% ARC to HOGG: Please suggest any nomenclature or definition changes as you see fit!
\begin{eqnarray}
	d_{ms} & = & \left(\frac{T_{{\rm eff},ms} - T_{{\rm eff},joint}}{\delta_{T_{{\rm eff},ms}}}\right)^2 + \left(\frac{\log{g}_{ms} - \log{g}_{joint}}{\delta_{\log{g},ms}}\right)^2   \nonumber \\
	d_{giant} & = & \left(\frac{T_{{\rm eff},giant} - T_{{\rm eff},joint}}{\delta_{T_{{\rm eff},giant}}}\right)^2 + \left(\frac{\log{g}_{giant} - \log{g}_{joint}}{\delta_{\log{g},giant}}\right)^2   \quad ,
\end{eqnarray}

\noindent{}as a metric to derive the weights,
\begin{eqnarray}
	w_{ms} = \frac{1}{{d_{ms}}^2} & \text{and} & w_{giant} = \frac{1}{{d_{giant}}^2} \quad ,
\end{eqnarray}

\noindent{}and produce the weighted labels $\hat\ell$:
\begin{eqnarray}
	\hat\ell = \frac{w_{ms}\ell_{ms} + w_{giant}\ell_{giant}}{w_{ms} + w_{giant}} \quad .
\end{eqnarray}

We calculate weighted errors of $\hat\ell$ and correlation coefficients in 
the same manner.  In Figure \ref{fig:model-weights} we show the mean relative
weight $w_{ms}/(w_{ms} + w_{giant})$ within each two-dimensional bin of 
$\hat\teff$ and $\hat\logg$.  For giant stars the relative weight of 
the main-sequence model is zero, and vice-versa for main-sequence stars.
The relative weights smoothly transition from 0 to 1 on the sub-giant branch
near $\log{g} \approx 3.5$, in the training set overlap region of both models.
For abundance labels in the giant model that are not in the main-sequence
model (e.g., [O/H], [Mg/H]), we only report abundances if there was no contribution
from the main-sequence model: $w_{ms} = 0$.

Our catalog now contains entries of $XXX$ stars with effective temperature $\teff$,
surface gravity $\logg$, iron abundance [Fe/H], and $XXX$ of those stars are giants
with elemental abundances of [O/H], [Al/H], [Mg/H], [Ca/H], [Si/H], and [Ni/H].
In total, this leads to a catalog of X.XX million stellar abundances. 
The weighted $\teff$ and $\logg$ values for stars meeting different
S/N constraints are shown in Figure \ref{fig:test-set-hrd}, both in logarithmic density
and mean metallicity.  The artefacts from individual models are no longer 
apparent, and the rich structure of the Hertzsprung-Russell is visible.

@@

In addition to the cross-validation tests that we have previously described, 
we have conducted a number of internal and external validation experiments to 
test the validity of our results.  We will begin by describing internal validation
tests based on repeat observations, before evaluating our accuracy based on
high-resolution literature comparisons.

@@

@@

\subsubsection{Comparison with \rave\ DR5}

Here we cross-match our results against literature samples to verify that
our results are accurate.  Because this work is an independent analysis, our
first point of reference is against the official \rave\ data release.  This
comparison is shown in Figure \ref{fig:rave-dr5-comparison}, 
where we show our results with respect to the fifth \rave\ data release \citep{Kunder_2016}.  In order
to provide a fair comparison, we only show the $XXX$ stars that meet a number of quality
flags in both samples: 
\stub{What selections qualify as a fair comparison in both samples?}
There is very good agreement in $\teff$, with a bias
and RMS of just XX~K and XXX~K, respectively.  We find our $\logg$ values are
closest to the `calibrated' \rave\ surface gravities, which make use of 
asteroseismic information to correct (increase) the $\logg$ values along the
giant branch.  The calibrated \rave\ surface gravities also correct the 
location of the red clump (by $\sim0.5$~dex) to be in the same position in our
results.  However, while our $\logg$ values generally agree with the calibrated
\rave\ values along the giant branch, our $\logg$ values do differ slightly 
along the main-sequence.  This is principally because the main-sequence in 
this work tends to taper down towards higher $\logg$ values at cooler 
temperatures, whereas the comparison sample tend to have a slightly flatter
main-sequence.  This difference is not likely to have a very significant 
effect on the detailed abundance or spectrophotometric distance determinations
between these studies \citep{Binney_2014}.

\subsubsection{Comparisons with Reddy, Bensby, and Valenti \& Fischer}

\stub{Caution re MP dwarfs appear as sub-giants in our sample}

\subsubsection{Comparison with the \ges\ survey}

@@

@@

After verifying that our stellar labels are comparable with other
high-resolution studies, here we verify that the abundances we find
are consistent with expectations from astrophysics.  One example is stars in
globular clusters, which show anti-correlations in light element abundances 
\citep[e.g.,][and references therein]{Norris_Da_Costa_1995,Carretta_2009}.  In the \rave\
survey, \cite{Anguiano_2015} found 70 stars with positions and radial velocities that are
consistent with being members of globular clusters: 49 stars belonging to NGC~5139 ($\omega$~Centauri), 11 members of the retrograde
globular cluster NGC~3201, and 10 members of NGC~362.  
All of these systems are well-studied and exhibit abundance anti-correlations 
in Na--O and Mg--Al \citep{Marino_2011,Carretta_2009,Carretta_2013,Munoz_2013}.  Other correlations are also present 
(e.g., C--N), but those relationships are less relevant for this work because 
we do not measure C and N abundances here.  We note that \emph{none} of these 
globular cluster stars are present in our training set.

@@

\acknowledgements
We thank 
	Jonathan Bird (Vanderbilt),
	Sven Buder (MPIA), 
and 
	Sergey Koposov (Cambridge).
This research made use of: 
  	NASA's Astrophysics Data System Bibliographic Services;
  	Astropy, a community-developed core Python package for Astronomy \citep{astropy};
and 
  	\project{TOPCAT} \citep{Taylor_2005}.
This work was partly supported by the European Union FP7 programme through ERC 
grant number 320360. Funding for K.~H.~ has been provided through the Simons 
Foundation Society of Fellows and the Marshall Scholarship.
Funding for RAVE has been provided by: the Australian Astronomical Observatory; 
the Leibniz-Institut fuer Astrophysik Potsdam (AIP); the Australian National 
University; the Australian Research Council; the French National Research Agency;
the German Research Foundation (SPP 1177 and SFB 881); the European Research 
Council (ERC-StG 240271 Galactica); the Istituto Nazionale di Astrofisica at 
Padova; The Johns Hopkins University; the National Science Foundation of the USA
(AST-0908326); the W. M. Keck foundation; the Macquarie University; the 
Netherlands Research School for Astronomy; the Natural Sciences and Engineering 
Research Council of Canada; the Slovenian Research Agency; the Swiss National 
Science Foundation; the Science \& Technology Facilities Council of the UK; 
Opticon; Strasbourg Observatory; and the Universities of Groningen, Heidelberg 
and Sydney. The RAVE web site is https://www.rave-survey.org.

@@

\bibitem[Carretta et al.(2009)]{Carretta_2009} Carretta, E., Bragaglia, A., Gratton, R.~G., et al.\ 2009, \aap, 505, 117

\bibitem[Carretta et al.(2013)]{Carretta_2013} Carretta, E., Bragaglia, A., Gratton, R.~G., et al.\ 2013, \aap, 557, A138

@@

\bibitem[Garc{\'{\i}}a P{\'e}rez et al.(2016)]{Garcia_Perez_2016} Garc{\'{\i}}a P{\'e}rez, A.~E., Allende Prieto, C., Holtzman, J.~A., et al.\ 2016, \aj, 151, 144

@@

\bibitem[Marino et al.(2011)]{Marino_2011} Marino, A.~F., Milone, A.~P., Piotto, G., et al.\ 2011, \apj, 731, 64

@@

\bibitem[Mu{\~n}oz et al.(2013)]{Munoz_2013} Mu{\~n}oz, C., Geisler, D., \& Villanova, S.\ 2013, \mnras, 433, 2006

@@

\bibitem[Norris \& Da Costa(1995)]{Norris_Da_Costa_1995} Norris, J.~E., \& Da Costa, G.~S.\ 1995, \apj, 447, 680

@@

\bibitem[Reddy et al.(2003)]{Reddy_2003} Reddy, B.~E., Tomkin, J., Lambert, D.~L., \& Allende Prieto, C.\ 2003, \mnras, 340, 304

\bibitem[Reddy et al.(2006)]{Reddy_2006} Reddy, B.~E., Lambert, D.~L., \& Allende Prieto, C.\ 2006, \mnras, 367, 1329

@@

\bibitem[Valenti \& Fischer(2005)]{Valenti_Fischer_2005} Valenti, J.~A., \& Fischer, D.~A.\ 2005, \apjs, 159, 141

@@

\begin{figure*}[p]
\includegraphics[width=\textwidth]{figures/test-set-density.pdf}
\caption{\label{fig:test-set-density}}
\end{figure*}

@@

\begin{figure}[p]
\center
\includegraphics[width=0.5\textwidth]{figures/joint-model-differences.pdf}
\caption{The normalized differences in effective temperature $\teff$ and surface gravity $\logg$ between the main-sequence model and the joint model (top panel), and the giant model and the joint model (bottom panel).  The density scaling is logarithmic, and the differences in $\teff$ and $\logg$ are scaled to make them approximately isotropic (see text for details).  The peak at $(0, 0)$ represents good agreement between the joint model and comparison model, whereas the over-densities elsewhere are a consequence of testing the model on stars very different to the training set (e.g., dwarf stars tested on a model trained with only giant stars).\label{fig:joint-model-differences}}
\end{figure}

\begin{figure}[p]
\center
\includegraphics[width=0.5\textwidth]{figures/model-weights.pdf}
\caption{The mean relative main-sequence model weight $w_{ms}/(w_{ms} + w_{giant})$ at each hexagonal bin of weighted effective temperature $\teff$ and surface gravity $\logg$.  The relative weighting illustrates how only results from the main-sequence model are adopted for unevolved stars, and there is a gradual transition to using results from the giant model, before only results from the giant model are used for evolved stars.\label{fig:model-weights}}
\end{figure}
